Inferring true changes in microbial abundance from taxonomically-biased microbiome measurements

### 2022-02-04 case-studies.Rmd
Revise fungal case study
Revised based on BJC's edits and comments on 2022-02-01.

Measurement of control communities along with the primary experimental samples can enable researchers to directly measure and remove the effect of bias prior to or concurrent with downstream DA analysis (@mclaren2019cons).
Gnotobiotic community experiments are well suited to this form of _calibration via community controls_ since, unlike most natural ecosystems, it is possible to assemble 'mock communities' containing all species in known relative abundances.

@leopold2020host are the first to use mock-based calibration in a gnotobiotic microbiome experiment.
In a study of host-commensal-pathogen interactions, @leopold2020host inoculated plants with 8 commensal fungal species and subsequently exposed plants to a fungal pathogen.
To investigate the joint effect of colonization order and host genotype, plants varied in source location and in which commensal colonized first.
The authors used ITS amplicon sequencing to measure communities before and after pathogen infection, along with mock communities with quantified genome concentrations of DNA from the 9 species.
These DNA mock communities allowed the authors to directly measure the bias due to the sequencing workflow following DNA extraction and correct for it when analyzing the primary experimental samples following the method of @mclaren2019cons.
Although extraction is not accounted for, considerable bias might still be expected due to preferential PCR binding and amplification and the substantial ribosomal copy-number-variation (CNV) in fungi (@lofgren2019geno).
<!-- TODO: add citations -->
The authors found a 13X difference between the most and least efficiently measured commensal, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.
We reanalyzed this experiment to understand the impact that bias might have had on the study's analyses were it not accounted for.

@leopold2020host used the ITS sequencing profiles of the pre-infection samples to understand the impact of host genotype and an experimental treatment---which species was allowed to colonize first---on the commensal community composition prior to infection.
Two DA analyses were conducted to address these questions.
First, the authors used linear regression of log species proportion (via negative-binomial regression to account for random read-sampling error) to ask whether host genotype, treatment, and their interaction significantly impacted overall community composition.
Second, the authors quantified the strength of priority effects---the advantage of being allowed to colonize first---in different host genotypes by the LFC in the proportion of a species when allowed to colonize first versus later with the majority of commensals.

Since these analysis are both based on proportions, bias could in principle impact their results.
We re-ran both analyses with and without bias correction, finding nearly identical results regardless of whether bias is accounted for.^[The two sets of analyses can be seen at the following links: [negative-binomial regression analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-06-leopold2020host-original-regression-analysis/index.html); [priority-effects analysis](https://mikemc.github.io/differential-abundance-theory/notebook/posts/2022-01-05-leopold2020host-priority-effects/index.html)]
To understand why, we compared the multiplicative variation taxa proportions with that of the mean efficiency.
The proportions of each taxon varied substantially across samples, much more so than the mean efficiency, which had a geometric range of 1.62X and a geometric standard deviation of just 1.05X (SI Figure \@ref(fig:leopold2020host-variation)).
<!-- Explanation -->
Why does the mean efficiency vary so little despite substantial multiplicative variation in the efficiency among taxa and in the taxa proportions across samples?
This answer relates to the fact that the mean efficiency is a weighted _arithmetic_ average of species proportions and so is insensitive to large multiplicative variation in low-proportion species.
The pre-infection samples in the @leopold2020host study are always dominated by the three species with the highest efficiencies, whose efficiencies vary only by 3X (and just 1.5X between the two most dominant).
The lowest efficiency species, despite having large variation in their proportion on the log scale, are always rare (proportions $\ll 1$, even after bias correction) and hence don't significantly affect the mean efficiency.
<!-- Conclusion -->
Because the log mean efficiency varies much less than the log proportions for each taxon, there is little room for it to distort the results of DA analyses based on variation in log proportions.

<!-- Notes: -->
<!-- - Consider whether to use log-transformed abundances and efficiencies throughout, and standard (non-geometric) ranges and standard deviations -->
<!-- - Consider connecting back to the previous section and reminding of the condition for the variation in mean efficiency to impact regression analyses -->
<!-- - Consider comparing the correlations among taxa with covariates; or at least looking at that of the mean efficiency, since correlation is also an important piece of the requirement -->
<!-- - The priority-effect analysis isn't done as a regression, though it could have been (but would have been a slightly different calculation). I might want to explain this more here and in the notebook post. -->
<!-- - I don't consider/mention whether there are actually 'significant' DA results; that would perhaps be a useful piece of the story. Or: I should perhaps say what the result was of each of the two analyses, so it is more concrete what results are not changing. -->
<!-- - Consider expanding on the impact of bias, using Penicillium as an example: It's proportion is generally ~6X lower without correction, but its changes across samples are not affected by correction -->

We additionally considered the potential for bias to impact analysis of how commensal taxa responded to infection.
In particular, it is interesting to consider whether any commensals are able to increase in absolute concentration in response to infection.
The experiment performed by @leopold2020host does not allow absolute-abundance measurement; nevertheless, the results of the previous sections allow us to consider the impact that bias would have on an analysis of absolute genome concentration measured using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)).
It is first useful to consider how bias impacts the estimated change in log proportion of each commensal following infection.
The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, raising the mean efficiency of post-infection samples (SI Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)).
Across different host genotypes, the average increase in mean efficiency ranges from 2.5X to 5.2X. 
We used simple linear regression to measure the average change in log proportion of each commensal for each host genotype, with and without bias correction.
Regardless of whether bias is accounted for, the log proportion of each commensal decreased, which is unsurprising given the pathogen's growth and the sum-to-one constraint of species proportions.
Without bias correction, however, the LFCs were lower by an amount corresponding to the inverse change in log mean efficiency.
The magnitude errors created were substantial for a large fraction of commensal-host pairs; in several cases, the LFC estimates without bias correction were clearly negative, whereas the estimates with bias correction were close to 0.

Now, consider the impact this shift in mean efficiency might have on a regression analysis with log absolute genome concentration as the response, if measured using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)).
We consider two scenarios.
_Scenario 1:_ Total genome concentration in each sample is perfectly known and used to measure the concentrations of individual taxa via Equation \@ref(eq:density-prop-meas).
In this case, bias in the MGS measurements would create absolute errors in the LFCs for genome concentration identical to those for proportions.
The scientific error, however, may be much worse.
Since the sum-to-one constraint of proportions no longer applies, the LFCs will necessarily be larger (less negative), such that the absolute error caused by the change in mean efficiency could plausibly create substantial biological errors where some taxa that persist or increase in abundance instead appear to decrease.
_Scenario 2:_ Total ITS concentration is measured by qPCR with the same primers and PCR protocol used for ITS sequencing, and used without any correction to measure individual taxa concentration with the total-abundance normalization method.
In this case, we should expect the increase in mean efficiency to have little to no impact on LFCs, since it would be offset by the increase in mean efficiency of the total-concentration measurement.
Moreover, since the total ITS concentration and ITS sequencing are both performed on the extracted DNA, we can also expect the effect of taxonomic bias caused by DNA sequencing to be offset.
Thus, the systematic shift in mean efficiency following infection may or may not significantly impact the LFC estimates depending on how the total-abundance measurement is made prior to normalization.

<!-- cut text -->
<!-- - Genome concentration was chosen as the measure of abundance, due to the tendency for fungi to grow in forms where cells are not easily demarcated. -->
<!-- - The original study focused on how host genotype and the pre-infection community affected disease severity after challenge with a pathogen, and did not analyze the impact of infection on commensals. -->
<!-- - The post-colonization samples were not analyzed by @leopold2020host in part because the reads from post-colonization samples tended to be dominated by the pathogen, giving little ability to infer the relative abundance of the rarer and lower efficiency species. -->
<!-- - This was presumably due to high pathogen growth and aided by the pathogen's post-extraction efficiency being substantially larger than the other species; the high efficiency of the pathogen typically leads to a roughly 6X increase in the mean efficiency of post-colonization samples (REF SI/FIG). -->

>>>>

The taxonomic bias species present in 'mock communities' of known composition can be directly measured, and the measured bias can then be used to correct downstream DA analysis (@mclaren2019cons).
In practice it is difficult to construct control communities that span the many species present in complex natural communities.
However, gnotobiotic community experiments are well suited to this form of _calibration via community controls_ since, unlike most natural ecosystems, it is possible to assemble mock communities containing all species in known relative abundances.

In a study of host-commensal-pathogen interactions, @leopold2020host inoculated plants with 8 commensal fungal species and subsequently exposed plants to a fungal pathogen.
The authors used ITS amplicon sequencing to measure communities before and after pathogen infection.
Motivated by the substantial ribosomal copy-number variation (CNV) in fungi (@lofgren2019geno), the authors also performed control measurements of mock communities that they constructed from quantified genomic DNA of the 9 species in the experiment; these controls were used to measure taxonomic bias with the method of @mclaren2019cons.
The authors found a 13X difference between the most and least efficiently measured commensal, while the pathogen was measured 40X more efficiently than the least efficiently measured commensal.

@leopold2020host performed two related DA analyses on the pre-infection communities: the first characterized the relative importance of host genetics and species arrival order on species relative abundances in the fully-established community, and the second quantified the strength of 'priority effects'---the advantage gain by a species from being allowed to colonize first.
Both analyses were based on fold changes in species proportions and so in principle were sensitive to taxonomic bias.
To ensure the results were accurate, the authors incorporated the bias measured from the control samples with analysis-specific calibration procedures.

Calibration had negligible impact on the results (personal communication with Devin Leopold and confirmed by our own reanalysis).
To understand why, we examined the variation in species proportions and the mean efficiency across the pre-infection communities (SI Figure \@ref(fig:leopold2020host-variation)).
Despite the 13X variation in the efficiencies among species, the mean efficiency hardly varied across samples (SI Figure \@ref(fig:leopold2020host-variation)C), having a geometric range of 1.62X and a geometric standard deviation of 1.05X.
This consistency in the mean efficiency was despite the fact that each species each showed substantial multiplicative variation (SI Figure \@ref(fig:leopold2020host-variation)A).
But the pre-infection samples were always dominated by the three species with the highest efficiencies, which varied by 3X and by just 1.5X between the two most dominant.
The mean efficiency, equal to the proportion-weighted arithmetic average of species efficiencies, is insensitive to species present at low relative abundance and so remained relatively constant across samples.
Because the multiplicative variation in the mean efficiency was much smaller than that in the proportions of individual taxa, it had a negligible impact on the inferred fold changes and the DA analyses based on them.

We performed an additional DA analysis on the data from @leopold2020host to investigate whether any commensals increased in absolute concentration in response to infection.
The pathogen is absent in pre-infection samples but tends to dominate the community post-infection, resulting in a substantially higher mean efficiency in post-infection samples (SI Figure \@ref(fig:leopold2020host-infection-mean-efficiency-dist)).
Across different host genotypes, the average post-infection increase in mean efficiency ranged from 2.5X to 5.2X.
We used simple linear regression to measure the LFC in commensal species proportions pre- to post-infection, with and without calibration (SI Figure \@ref(fig:leopold2020host-infection-lfc)).
All commensals tended to decrease in proportion post-infection, which is expected given the pathogen’s growth to most abundant community member and the sum-to-one constraint of species proportions.
However, when taxonomic bias was ignored the magnitude of the post-infection decrease was overestimated, by an amount corresponding to the inverse change in log mean efficiency.
This magnitude error was substantial for the commensal-host pairs with the weakest changes; in several cases, LFC estimates with 95% confidence intervals below 0 became indistinguishable from 0 after calibration.

Although @leopold2020host did not include absolute-abundance measurements, we can consider the impact taxonomic bias would have on an absolute DA analysis in a simple scenario in which total genome concentration of each pre- and post-infection sample is perfectly known.
The total genomic concentrations of each sample, combined with the species proportions measured by MGS, can be combined into a measurement of species concentrations using the total-abundance normalization method (Equation \@ref(eq:density-prop-meas)).
In this case, the bias in the MGS measurements will create absolute errors in the estimated LFCs of genome concentration of equal magnitude to the errors in the LFC estimates for proportions.
The scientific error, however, might become worse.
Suppose that the species whose calibrated proportions appeared to hold steady or slightly decrease actually increased in absolute abundance by around 2X.
As bias shifts our estimates downwards by 2.5X to 5.2X, we would instead conclude that these taxa decreased.
